A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-19 tex/codatadp.tex
step -> phase

\paragraph{Step 1: Computing the Bidirectional Closure.}
Figure~\ref{fig:cc-rules} gives the basic rules of the calculus.
Together with \rn{Refl}, \rn{Sym}, and \rn{Trans}, the \rn{Cong} rule computes the (upward) congruence closure,
whereas the \rn{Decomp} and \rn{Clash} rules together compute (downward) unification.
In the unification rules, additional equalities are inferred based on the injectivity of constructors by \rn{Decomp},
and failures to unify equated terms are recognized by \rn{Clash}.
The rule \rn{Conflict} recognizes when an equality and its negation both occur in $\Ec$, in which case $\Ec$ has no model.

@@

\paragraph{Step 2: Applying Acyclicity and Bisimilarity.}
In the second step,
we assume that $\Ec$ is saturated with respect to the rules in Figure~\ref{fig:cc-rules}---that is,
no rule can be applied to extend~$\Ec$ further.
When this is the case, it is easy to see that $\Ec$ induces an equivalence relation over $\tEc$ such that two terms $t_1$ and $t_2$ are equivalent if and only if $t_1 \teq t_2 \in \Ec$.
Thus, we can consider $\Ec$ as a set of equivalence classes of terms. 
For a term $t \in \tEc$, we write $\ec{t}$ to denote the equivalence class in $\Ec$ that contains $t$.

%For presentation of the rules in this step, we rely on a representation of terms in the $\mu$-notation.
The rules of this step contain premises that involve a mapping $\Val$ from equivalence classes to terms containing $\mu$-bindings.
Recall that $\mu$-bindings can be used for representing possibly cyclic terms and values.
%(Section~\ref{sec:introduction}).
% for example, $\mu x. \const{C}( \const{0}, x )$ represents the cyclic value $\const{C}( \const{0}, \const{C}( \const{0}, \ldots ))$.
Formally, $\mu$-terms are defined recursively as being either a variable $x$
or a node of the form
$\mu x.\; \const{C}( \vec t\, )$ for some constructor $\const{C} \in \Ctr$ and
$\mu$-terms $\vec t$ with the expected types.
The variable $x$ need not occur under the binder's body, in which case the $\mu$-prefix
is optional.

@@

Intuitively, a $\mu$-term $t$ of type $\tau$ describes a class of values of $\tau$, 
%either in the case when $\tau$ is a codatatype or datatype type,
where a cyclic $t$ describes an infeasible class of values when
%$\tau \not\in \Codata$.
$\tau \in \Data$.
At the beginning of this step, 
we compute a mapping $\Val$ from equivalence classes to $\mu$-terms 
that describe the class of values that terms in that equivalence class can take in the models of $\Ec$. 
For each equivalence class $\ec{x}$, we associate a fresh variable $v_{\ec{x}}$ of the same type as $x$ not occurring in $\tEc$,
and initially say that $\Val$ contains $\ec{x} \mapsto v_{\ec{x}}$,
in other words, there are initially no constraints on the values for any equivalence class $\ec{x}$.
Then, $\Val$ is modified by applying the following update exhaustively:
\[
\inferrule{
  v_{\ec{x}} \in FV( \Val )
  \quad
  \const{C}( t_1, \ldots, t_n ) \in \ec{x}
  \quad
  \const{C} \in \Ctr
}{
  \Val := \Val \{ v_{\ec{x}} \mapsto \mu v_{\ec{x}}. \const{C}( v_{\ec{t_1}}, \ldots, v_{\ec{t_n}} ) \}
}
\]
Here, we write $FV( \Val )$ to denote the \emph{free variables} occurring in $\Val$, which
excludes variables that occur beneath $\mu$-binders.
It is straightforward to see that the height of terms produced as a result of this update
is bounded by the number of equivalence classes of $\Ec$,
and thus our construction of $\Val$ is terminating.

@@

\begin{example}
Let $\const{C}$ be a constructor of type $\tau_1 \times \tau_2 \rightarrow \tau_2$,
and let $\Ec$ be the set $\{ x \teq \const{C}(y,\const{C}(y,x)) \}$.
After applying the rules in Figure~\ref{fig:cc-rules} to completion, the equivalence classes of $\Ec$ are
$\{ x, \const{C}(y,\const{C}(y,x)) \}$, $\{ \const{C}(y,x) \}$, and $\{ y \}$.
We then construct $\Val$ using the aforementioned steps, after which we obtain:
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}r@{\hspace{.5em}}l@{\hspace{.5em}}r@{\hspace{.5em}}l}
\Val \ec{x} & = & 
\mu v_{\ec{x}}. & \const{C}( v_{\ec{y}}, & \mu v_{\ec{\const{C}(y,x)}}. & \const{C}( v_{\ec{y}}, v_{\ec{x}} ) ) \\
\Val \ec{\const{C}(y,x)} & = & 
\mu v_{\ec{\const{C}(y,x)}}. & \const{C}( v_{\ec{y}}, & \mu v_{\ec{x}}. & \const{C}( v_{\ec{y}}, v_{\ec{\const{C}(y,x)}} ) )
\end{array}\]
Notice that $\Val \ec{x}$ and $\Val \ec{\const{C}(y,x)}$ are syntactically equivalent for the renaming
$\{ v_{\ec{x}} \mapsto v_{\ec{\const{C}(y,x)}}, v_{\ec{\const{C}(y,x)}} \mapsto v_{\ec{x}} \}$,
and thus our calculus may infer $x \teq \const{C}(y,x)$.
Intuitively, this equality holds, since the values of $x$ and $\const{C}( y, x )$ 
are equivalent for any value of $v_{\ec{y}}$.
\qed
\end{example}

@@

\paragraph{Step 3: Branching.}
If neither a rule from Figure \ref{fig:cc-rules}~or~\ref{fig:ab-rules} applies to $\Ec$, 
the rules from Figure~\ref{fig:split-rule} are applied.
The rule \rn{Singleton} handles the case when there are two terms from $\tEc$ having a co-recursive singleton type,
in which case they must be equivalent.
Otherwise, in the case there is a selector term applied to $t$, or if $t$ has a finite datatype type, 
then the equivalence class of $t$ must contain an application of one of the constructors of $\tau$ (where $\Ctr^\tau = \{ C_1, \ldots C_n \}$), 
which is enforced by the rule \rn{Split}.

@@

In the following, we consider only normal models.
Thus, when constructing a model $\M$ for $\Ec$,
it remains only to define how $\M$ interprets wrong-applied selector terms and variables.
For the latter, this will be based on the mapping $\Val$ constructed in step two of our calculus.
We compute the \emph{completion} of $\Val$ for normal model $\M$, which we write as $\Val^\ast$,
which we construct by assigning values from $\M$ to unassigned variables in the domain of $\Val$.

@@

\begin{lemma}[Solution Soundness]%
\label{lem:ss}%
\afterDot
If there exists a derivation tree with root node $\Ec$ containing a saturated node, then $\Ec$ is satisfiable in $\thD$.
\end{lemma}
\begin{proof}
Let $\Fc$ be a saturated node in a derivation tree with root node $\Ec$.
%We will construct a model $\M$ for a set of equalities $\Fc$ that is equivalent to $\Fc_0$, and where 
%all equivalence classes of $\Fc$ contain at least one variable, and all selectors in $\Fc$ are applied to variables only.
%The former comes with no loss of generality since new equalities of the form $y \teq t$ for fresh variable $y$ can be added to $\Fc$ without affecting its satisfiability,
%The latter also comes with no loss of generality since nested applications of selectors in $\Fc_0$ can be replaced by fresh variables while adding additional equalities to $\Fc$.
We consider a normal model $\M$ 
that interprets wrongly-applied selectors based on equality information in $\Fc$,
and interprets the variables of $\Fc$ based on the completion $\Val^\ast$ of the mapping $\Val$ from step two of our calculus.
For the latter, let $\M( x )$ be the value in $\M$ that is bisimilar with $\Val^\ast \ec{x}$ for each variable $x$ in $\mathcal{T}(\Fc)$.

@@

Discriminators are treated as predicate symbols, so a discriminator $\const{d}_i( t )$ is such that $\const{d}_i$ 
is a predicate symbol, instead of reducing this constraint to $t \teq \const{C}_i( s^1_i( t ), \ldots, s^n_i( t ) )$.
We have found that this leads to better performance since this reduction introduces terms more eagerly to $\tEc$.
Handling discriminators requires extending the decision procedure with several additional rules (see~\cite{} for more details), which apply uniformly to both datatypes and codatatypes.
Also, selectors are collapsed eagerly:
if $s^j_i( t )$ is a term in $\tEc$ and $t$ contains the constructor 
$\const{C}_i( \vec u )$, then we immediately infer $s^j_i( t ) \teq u_j$, whereas the calculus we presented would apply \rn{Split} and \rn{Decomp} before inferring this equality.  
For the sake of reducing the number of unique constraints considered by the calculus, we compute a normal form for constraints as a pre-processing step.
In particular, we 
replace $t \teq s$ by $s \teq t$ if $s$ is less than $t$ based on a term ordering,
replace $\const{C}_i( \vec t ) \teq \const{C}_j( \vec u )$ with $\bot$ when $i \neq j$,
replace all selector terms of the form $\const{s}^{\,i}_j( \const{C}_j(t_1,\ldots,t_n) )$ by $t_i$,
and replace all occurrences of discriminators $\const{d}_i( \const{C}_k( \vec t ) )$ by $\top$ if $i=k$ and $\bot$ if $i \neq k$.

@@

Step 1 of the calculus (along with the optimization for collapsing selectors) and the \rn{Singleton} rule are performed eagerly, 
that is for partial satisfying assignments $M$, while step 2 and the \rn{Split} rule from step 3 are performed only for complete satisfying assignments $M$, and in that order.

>>>>

\paragraph{Phase 1: Computing the Bidirectional Closure.}
Figure~\ref{fig:cc-rules} gives the basic rules of the calculus.
Together with \rn{Refl}, \rn{Sym}, and \rn{Trans}, the \rn{Cong} rule computes the (upward) congruence closure,
whereas the \rn{Decomp} and \rn{Clash} rules together compute (downward) unification.
In the unification rules, additional equalities are inferred based on the injectivity of constructors by \rn{Decomp},
and failures to unify equated terms are recognized by \rn{Clash}.
The rule \rn{Conflict} recognizes when an equality and its negation both occur in $\Ec$, in which case $\Ec$ has no model.

@@

\paragraph{Phase 2: Applying Acyclicity and Bisimilarity.}
In the second phase,
we assume that $\Ec$ is saturated with respect to the rules in Figure~\ref{fig:cc-rules}---that is,
no rule can be applied to extend~$\Ec$ further.
When this is the case, it is easy to see that $\Ec$ induces an equivalence relation over $\tEc$ such that two terms $t_1$ and $t_2$ are equivalent if and only if $t_1 \teq t_2 \in \Ec$.
Thus, we can consider $\Ec$ as a set of equivalence classes of terms. 
For a term $t \in \tEc$, we write $\ec{t}$ to denote the equivalence class in $\Ec$ that contains $t$.

%For presentation of the rules in this phase, we rely on a representation of terms in the $\mu$-notation.
The rules of this phase contain premises that involve a mapping $\Val$ from equivalence classes to terms containing $\mu$-bindings.
Recall that $\mu$-bindings can be used for representing possibly cyclic terms and values.
%(Section~\ref{sec:introduction}).
% for example, $\mu x. \const{C}( \const{0}, x )$ represents the cyclic value $\const{C}( \const{0}, \const{C}( \const{0}, \ldots ))$.
Formally, $\mu$-terms are defined recursively as being either a variable $x$
or a node of the form
$\mu x.\; \const{C}( \vec t\, )$ for some constructor $\const{C} \in \Ctr$ and
$\mu$-terms $\vec t$ with the expected types.
The variable $x$ need not occur under the binder's body, in which case the $\mu$-prefix
is optional.

@@

Intuitively, a $\mu$-term $t$ of type $\tau$ describes a class of values of $\tau$, 
%either in the case when $\tau$ is a codatatype or datatype type,
where a cyclic $t$ describes an infeasible class of values when
%$\tau \not\in \Codata$.
$\tau \in \Data$.
At the beginning of this phase, 
we compute a mapping $\Val$ from equivalence classes to $\mu$-terms 
that describe the class of values that terms in that equivalence class can take in the models of $\Ec$. 
For each equivalence class $\ec{x}$, we associate a fresh variable $v_{\ec{x}}$ of the same type as $x$ not occurring in $\tEc$,
and initially say that $\Val$ contains $\ec{x} \mapsto v_{\ec{x}}$,
in other words, there are initially no constraints on the values for any equivalence class $\ec{x}$.
Then, $\Val$ is modified by applying the following update exhaustively:
\[
\inferrule{
  v_{\ec{x}} \in FV( \Val )
  \quad
  \const{C}( t_1, \ldots, t_n ) \in \ec{x}
  \quad
  \const{C} \in \Ctr
}{
  \Val := \Val \{ v_{\ec{x}} \mapsto \mu v_{\ec{x}}. \const{C}( v_{\ec{t_1}}, \ldots, v_{\ec{t_n}} ) \}
}
\]
Here, we write $FV( \Val )$ to denote the \emph{free variables} occurring in $\Val$, which
excludes variables that occur beneath $\mu$-binders.
It is straightforward to see that the height of terms produced as a result of this update
is bounded by the number of equivalence classes of $\Ec$,
and thus our construction of $\Val$ is terminating.

@@

\begin{example}
Let $\const{C}$ be a constructor of type $\tau_1 \times \tau_2 \rightarrow \tau_2$,
and let $\Ec$ be the set $\{ x \teq \const{C}(y,\const{C}(y,x)) \}$.
After applying the rules in Figure~\ref{fig:cc-rules} to completion, the equivalence classes of $\Ec$ are
$\{ x, \const{C}(y,\const{C}(y,x)) \}$, $\{ \const{C}(y,x) \}$, and $\{ y \}$.
We then construct $\Val$ using the aforementioned phases, after which we obtain:
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}r@{\hspace{.5em}}l@{\hspace{.5em}}r@{\hspace{.5em}}l}
\Val \ec{x} & = & 
\mu v_{\ec{x}}. & \const{C}( v_{\ec{y}}, & \mu v_{\ec{\const{C}(y,x)}}. & \const{C}( v_{\ec{y}}, v_{\ec{x}} ) ) \\
\Val \ec{\const{C}(y,x)} & = & 
\mu v_{\ec{\const{C}(y,x)}}. & \const{C}( v_{\ec{y}}, & \mu v_{\ec{x}}. & \const{C}( v_{\ec{y}}, v_{\ec{\const{C}(y,x)}} ) )
\end{array}\]
Notice that $\Val \ec{x}$ and $\Val \ec{\const{C}(y,x)}$ are syntactically equivalent for the renaming
$\{ v_{\ec{x}} \mapsto v_{\ec{\const{C}(y,x)}}, v_{\ec{\const{C}(y,x)}} \mapsto v_{\ec{x}} \}$,
and thus our calculus may infer $x \teq \const{C}(y,x)$.
Intuitively, this equality holds, since the values of $x$ and $\const{C}( y, x )$ 
are equivalent for any value of $v_{\ec{y}}$.
\qed
\end{example}

@@

\paragraph{Phase 3: Branching.}
If neither a rule from Figure \ref{fig:cc-rules}~or~\ref{fig:ab-rules} applies to $\Ec$, 
the rules from Figure~\ref{fig:split-rule} are applied.
The rule \rn{Singleton} handles the case when there are two terms from $\tEc$ having a co-recursive singleton type,
in which case they must be equivalent.
Otherwise, in the case there is a selector term applied to $t$, or if $t$ has a finite datatype type, 
then the equivalence class of $t$ must contain an application of one of the constructors of $\tau$ (where $\Ctr^\tau = \{ C_1, \ldots C_n \}$), 
which is enforced by the rule \rn{Split}.

@@

In the following, we consider only normal models.
Thus, when constructing a model $\M$ for $\Ec$,
it remains only to define how $\M$ interprets wrong-applied selector terms and variables.
For the latter, this will be based on the mapping $\Val$ constructed in the second phase of our calculus.
We compute the \emph{completion} of $\Val$ for normal model $\M$, which we write as $\Val^\ast$,
which we construct by assigning values from $\M$ to unassigned variables in the domain of $\Val$.

@@

\begin{lemma}[Solution Soundness]%
\label{lem:ss}%
\afterDot
If there exists a derivation tree with root node $\Ec$ containing a saturated node, then $\Ec$ is satisfiable in $\thD$.
\end{lemma}
\begin{proof}
Let $\Fc$ be a saturated node in a derivation tree with root node $\Ec$.
%We will construct a model $\M$ for a set of equalities $\Fc$ that is equivalent to $\Fc_0$, and where 
%all equivalence classes of $\Fc$ contain at least one variable, and all selectors in $\Fc$ are applied to variables only.
%The former comes with no loss of generality since new equalities of the form $y \teq t$ for fresh variable $y$ can be added to $\Fc$ without affecting its satisfiability,
%The latter also comes with no loss of generality since nested applications of selectors in $\Fc_0$ can be replaced by fresh variables while adding additional equalities to $\Fc$.
We consider a normal model $\M$ 
that interprets wrongly-applied selectors based on equality information in $\Fc$,
and interprets the variables of $\Fc$ based on the completion $\Val^\ast$ of the mapping $\Val$ from the second phase of our calculus.
For the latter, let $\M( x )$ be the value in $\M$ that is bisimilar with $\Val^\ast \ec{x}$ for each variable $x$ in $\mathcal{T}(\Fc)$.

@@

Discriminators are treated as predicate symbols, so a discriminator $\const{d}_i( t )$ is such that $\const{d}_i$ 
is a predicate symbol, instead of reducing this constraint to $t \teq \const{C}_i( s^1_i( t ), \ldots, s^n_i( t ) )$.
We have found that this leads to better performance since this reduction introduces terms more eagerly to $\tEc$.
Handling discriminators requires extending the decision procedure with several additional rules (see~\cite{} for more details), which apply uniformly to both datatypes and codatatypes.
Also, selectors are collapsed eagerly:
if $s^j_i( t )$ is a term in $\tEc$ and $t$ contains the constructor 
$\const{C}_i( \vec u )$, then we immediately infer $s^j_i( t ) \teq u_j$, whereas the calculus we presented would apply \rn{Split} and \rn{Decomp} before inferring this equality.  
For the sake of reducing the number of unique constraints considered by the calculus, we compute a normal form for constraints as a preprocessing step.
In particular, we 
replace $t \teq s$ by $s \teq t$ if $s$ is less than $t$ based on a term ordering,
replace $\const{C}_i( \vec t ) \teq \const{C}_j( \vec u )$ with $\bot$ when $i \neq j$,
replace all selector terms of the form $\const{s}^{\,i}_j( \const{C}_j(t_1,\ldots,t_n) )$ by $t_i$,
and replace all occurrences of discriminators $\const{d}_i( \const{C}_k( \vec t ) )$ by $\top$ if $i=k$ and $\bot$ if $i \neq k$.

@@

Phase 1 (along with the optimization for collapsing selectors) and the \rn{Singleton} rule are performed eagerly, 
that is for partial satisfying assignments $M$, while phase~2 and the \rn{Split} rule from phase~3 are performed only for complete satisfying assignments $M$, and in that order.
